Computing MEM seeding probabilities

### 2019-02-18 MEM.tex
Improved the abstract.

\begin{abstract}
High throughput sequencing produces massive amounts of data that must be
processed efficiently. Most reads are identified by mapping them to a
reference, typically a genome. Modern mappers rely on heuristics to gain
speed at a reasonable cost of accuracy. MEM seeding (where MEM stands for
Maximal Exact Match) is a popular heuristic that shows good empirical
performance but for which no theory is available. It is thus difficult to
calibrate or optimize mapping strategies based on MEM seeding. Here we
develop a theory to predict the performance of MEM seeding. More
specifically, we design a method to compute the probability a read is
mapped to the right location when using MEM seeds. Our computational
strategy is based on the analytic combinatorics programme, but uses only
part of it. By recoding the reads in a specialized combinatorial alphabet,
we can implicitly specify their generating function and extract the
probabilities of interest by direct methods or by Monte Carlo sampling.
This yields an efficient algorithm to calibrate the MEM seeding heuristic
for short read technologies.
\end{abstract}

@@

The focus of this document is not to solve the true location problem
\textit{per se}, but rather to evaluate whether a given solution is
correct. At first sight, whether we can map a read to its true location
seems to depend only on its length and on the error rate of the
sequencer. However, there is another important factor to consider: most
genomes contain repeated sequences, \textit{i.e.} relatively large
subsequences that are present at more than one location. For instance, we
cannot map the fragment with certainty if there exists an exact copy of
the sequence somewhere else in the genome, because we cannot know which of
the two copies was sequenced.

@@

We will refer to this task as the \emph{best location problem}. It amounts
to finding the optimal alignment between two sequences, and for this
reason has received substantial attention in bioinformatics. Finding the
best location of the read is the only way to get to its true location, so
we need to focus our attention on the best location problem.

@@

The most popular heuristic for this problem is to first identify short
regions of high similiarity between the read and the genome, and then run
an exact alignment algorithm around those regions. The first step is a
filtration called ``seeding'' and the regions of high similarity
are called ``seeds''. This is for instance the general strategy of the
popular alignment algorithm BLAST~\cite{pmid2231712}.

If seeds can be discovered fast, we can quickly zoom in on a small set of
candidate locations and reduce the input size of the (slow) alignment
algorithm. The disadvantage is that the target location is not always in
the candidate set. In this case, it cannot be discovered and the outcome
is either $ii.$ or $iii.$

This imposes a trade-off between speed and accuracy because each candidate
location must be tested with a slow algorithm. If the seeding step is set
to produce many candidates, the target is likely to be among them and it
will be discovered. The downside is that this will take long. Conversely,
if the seeding step is set to produce few candidates, the process will
run faster but the target is less likely to be in the candidate list so it
is less likely to be discovered.

Importantly, the terms of this trade-off depend on the algorithm. Some
seeding methods are better than others (see~\cite{sun2005designing,
pmid11934743,xu2006optimizing,pmid17044164,brejova2003vector,
pmid18684737} for examples of design, see~\cite{pmid16533404,pmid20460430}
for comparisons and discussions, and see~\cite{navarro2001guided} for a
more general review of filtration methods in pattern matching). The upshot
is that one can develop faster mapping algorithms at no cost for accuracy.

The seeding framework was developed for the best location problem, but it
also applies to the true location problem. Actually, the process is
exactly the same except that outcomes $i.$ and $ii.$ now refer to the true
location. Outcome $i.$, called a ``true positive'', is that the DNA
fragment is mapped to the \emph{true} location. Outcome $ii.$, called a
``false positive'' is that the DNA fragment is mapped to an incorrect
location. Outcome $iii.$, called a ``false negative'', does not change and
still means that the DNA fragment is not mapped at all.

We will make two assumptions that will simplify the discussion. The first
is that the true location is always the best. We will further comment on
this and explain how to relax this assumption in section
section~\ref{sec:not_the_best}. Until then, we consider that the true
location problem is the same as the best location problem. The second
assumption is that the best location is always discovered if it is in the
candidate set. This implies that outcomes $i.$, $ii.$ and $iii.$ are
determined by the seeding process. For this reason, we will sometimes
refer to them as the outcomes of the seeding step, even though they are
the outcomes of the whole mapping process.

@@

In general, seeds do not have to be exact matches~\cite{pmid20460430}. For
instance, the algorithm PatternHunter~\cite{pmid11934743} uses ``spaced
seeds'' that tolerate mismatches\footnote{For the anecdote, the method is
covered by a patent (reference US20070088510A1).}. However, in this
document, we will always assume that a seed is an exact match with no gap.
In other words, a seed is a perfect match between a part of the read and a
genomic site.

@@

\begin{definition}
A \emph{strict} MEM seed has a single genomic match.
A \emph{shared} MEM seed has several genomic matches.
\end{definition}

@@

@@

We have devised a method to compute the probability that the MEM seeding
heuristic misguides the mapping algorithm so that it cannot find the true
location of the read. The pillars of our strategy are borrowed from
analytic combinatorics~\cite{flajolet2009analytic,
sedgewick2013introduction}, even though we do not follow the standard
programme. Constructing generating functions usually serves the purpose of
finding their singularities in order to approximate the solution. In our
case, however, the weighted generating functions cannot be computed so
this strategy is not applicable.

>>>>

\begin{abstract}
The increasing throughput of DNA sequencing technologies demands faster
algorithms to analyze sequencing data. The fate of most reads is to be
mapped to a reference sequence, typically a genome. Modern mappers rely on
heuristics to gain speed at a reasonable cost for accuracy. MEM seeding
(where MEM stands for Maximal Exact Match) is a popular heuristic that
shows good empirical performance but for which theory is lacking. It is
thus difficult to calibrate or optimize mapping strategies based on MEM
seeding. Here we develop a theory to estimate the probability that the MEM
seeding heuristic fails to return the correct candidate. The major
innovation is to use concepts from analytic combinatorics to represent
reads as abstract sequences, from which we estimate the probabilities of
interest. We provide two separate methods, which combined together give a
practicial solution to the problem of calibrating MEM seeding when
aligning short reads. These results have implications for MEM-based
heuristics and lay the foundation of a general strategy to tackle sequence
alignment problems.
\end{abstract}

@@

At first sight, whether a read read can be mapepd to its correct location
seems to depend only on its length and on the error rate of the sequencer.
However, there is another important factor to consider: most genomes
contain repeated sequences, \textit{i.e.} relatively large subsequences
that are present at more than one location. For instance, we cannot map
the fragment with certainty if there exists an exact copy of the sequence
somewhere else in the genome, because we cannot know which of the two
copies was sequenced.

@@

We will refer to this task as the \emph{best location problem}. It amounts
to finding the optimal alignment between two sequences, and for this
reason has received substantial attention in bioinformatics. It turns out
that finding the best location of the read is the only way to get to its
true location, so for now we need to focus our attention on the best
location problem.

@@

The most popular heuristic for this problem is to first identify short
regions of high similiarity between the read and the genome, and then run
an exact alignment algorithm locally at those regions. The first step is a
filtration called ``seeding'' and the regions of high similarity are
called ``seeds''. This is for instance the general strategy of the popular
alignment algorithm BLAST~\cite{pmid2231712}.

If seeds can be discovered fast, we can quickly zoom in on a small set of
candidate locations and reduce the input size of the (slow) alignment
algorithm. The disadvantage is that the target location is not always in
the candidate set. In this case, it cannot be discovered and the outcome
is either $ii.$ or $iii.$, which are both erroneous.

The slowness of the alignment algorithm imposes a trade-off between speed
and accuracy. If the seeding step is set to produce many candidates, the
target is likely to be among them and it will be discovered. The downside
is that this will take long. Conversely, if the seeding step is set to
produce few candidates, the process will run faster but the target is less
likely to be in the candidate list so it is less likely to be discovered.

Importantly, the terms of this trade-off depend on the algorithm. Some
seeding methods are better than others (see~\cite{pmid20460430,
sun2005designing,pmid11934743,xu2006optimizing,pmid17044164,
brejova2003vector,pmid18684737,pmid15359419} for examples of design,
see~\cite{pmid16533404, pmid20460430} for comparisons and discussions, and
see~\cite{navarro2001guided} for a more general review of filtration
methods in pattern matching). The upshot is that we can develop faster
mapping algorithms at no cost for accuracy if we find better seeding
strategies.

The seeding framework was developed for the best location problem, but it
also applies to the true location problem. Actually, the process is
exactly the same except that outcomes $i.$ and $ii.$ now refer to the true
location. In outcome $i.$ called a ``true positive'', the DNA fragment is
mapped to the \emph{true} location. In outcome $ii.$ called a ``false
positive'', the DNA fragment is mapped to an incorrect location. Outcome
$iii.$ called a ``false negative'' is the same as before and means that
the DNA fragment is not mapped at all.

We will make two assumptions that will simplify the discussion. The first
is that the true location is always the best. We will further comment on
this and explain how to relax this assumption in section
section~\ref{sec:not_the_best}. Until then, we consider that the true
location problem is the same as the best location problem. The second
assumption is that the best location is always discovered if it is in the
candidate set. This implies that outcomes $i.$, $ii.$ and $iii.$ are
determined by the seeding process. For this reason, we will sometimes
refer to them as the outcomes of the seeding step, even though strictly
speaking they are the outcomes of the whole mapping pipeline\footnote{This
can be confusing because some authors refer to the off-target seeds as
false positives. When we say that the outcome of the seeding step is a
false positive, we mean that \emph{all} the seeds are off-target.}.

@@

In general, seeds do not have to be exact matches{}. For
instance, the algorithm PatternHunter~\cite{pmid11934743} uses ``spaced
seeds'' that tolerate mismatches\footnote{For the anecdote, the method is
covered by a patent (reference US20070088510A1).}. However, in this
document, we will always assume that a seed is an exact match with no gap.
In other words, a seed is a perfect match between a part of the read and a
genomic site.

@@

\begin{definition}
A \emph{strict} MEM seed has a single match in the genome.
A \emph{shared} MEM seed has several matches in the genome.
\end{definition}

@@

Algorithm~\ref{alg:mcmc} is a major result. It gives a compact solution to
the problem of estimating the probability that a read can be mapped when
using MEM seeds. The algorithm is also much faster than the naive approach
of sampling every nucleotide of every sequence, because it samples many
nucleotidea from \emph{all} the sequences at each iteration.

@@

We have devised a method to compute the probability that the MEM seeding
heuristic fails and sets up the mapping process for a false positive. This
fills a gap of theoretical knowledge to understand and calibrate the
performance of the MEM seeding heuristic. The pillars of our strategy are
borrowed from analytic combinatorics~\cite{flajolet2009analytic,
sedgewick2013introduction}, even though we do not follow the standard
programme. Constructing generating functions usually serves the purpose of
finding their singularities in order to approximate the solution. In our
case, however, the weighted generating functions cannot be computed so
this strategy is not applicable.
